\stktokens{}: Enforcing Well-Bracketed Control Flow and Stack Encapsulation Using Linear Capabilities

### 2020-07-24 popl-2019-journal-v/discussion.tex
Introduction for the discussion section.
A few final pre-submission modifications.

@@

  We have not investigated support for continuations and exceptions in \stktokens{} thoroughly but we expect such support could be added.
  For exceptions, one approach would let callers provide callees with an additional capability for exceptional returns.
  This second capability would be similar to the code part of the return capability pair and signed with the same seal.
  The callee would be able to invoke it to signal that an exception has been thrown after which the caller's code would handle the exception, either by executing an exception handler or by unwinding its stack frame and passing the exception on to its own caller.
  Essentially, this would mean every function would be made responsible for unwinding its own stack frame.
  Continuations are more complicated but could perhaps be treated using similar ideas.
  Alternatively, it might also be possible to have a piece of central trusted code that does the stack unravelling for all stack frames.
  To do this, this the trusted code would need to receive a copy of all return seals from the linker.

  % Say something about full-abstractions proofs and feature extensions
  Full-abstraction is a property of the whole language.
  In other words, a full-abstraction proof must consider all features of a language to make sure that the features don't interact in a way that breaks language abstractions.
  If we have a fully abstract compiler and add a feature to the source and target language, then the new compiler is no necessarily fully-abstract.
  Full abstraction would have to be proven for the new compiler to make sure that the new feature does not break existing abstractions in the language.
  Our proof of full-abstraction for \stktokens{} targets a simple capability machine that may not be able to enforce the high-level language abstractions we want, e.g.\ address hiding.
  In other words, the full-abstraction proof cannot be reused immediately.
  However, \stktokens{} is still a good candidate for the enforcement mechanism for well-bracketed control-flow and local-state encapsulation in a real fully abstract compiler.
  Generally speaking, it is worth investigating enforcement mechanisms for full abstraction in a simple setting that allows to quickly try out ideas and verify that the enforcement works.

  % new paragraph: conditional full-abstraction
  \dominique{This paragraph should perhaps also point out the link to the dynamic compromise paper?}
  \dominique{Then again, we come back to this later in the related work, so perhaps not necessary to discuss this here?}
  Our full-abstraction theorem, Theorem~\ref{thm:full-abstraction}, is not pure full-abstraction as it requires the components to be reasonable and well-formed.
  In other words, if we were to define a compiler phase that targets \srccm{}, then we would also have to show that every program it generates is well-formed and reasonable in order to use the full-abstraction result.
  Without the reasonability constraint, \stktokens{} would have to enforce reasonability instead.
  That is, \stktokens{} would have to dynamically ensure that no return seals or means to get return seals are passed in calls.
  Essentially, such checks would protect the trusted code against itself which shouldn't be necessary in the first place.
  Instead, the compiler should generate reasonable code that never exhibits the unreasonable behaviour.
  This should be done in a compiler phase where more information about the original program is available, e.g.\ the compilation phase that commits to the low-level translation.
  Similarly for the syntactic constraints given by well-formedness.
  The compiler should make sure to generate code that satisfies the well-formedness judgement, so it can be executed by the machine.

@@

Although this assumption can only be discharged by demonstrating an actual implementation with efficiency measurements, the following notes are based on private discussions with people from the CHERI team as well as our own thoughts on the matter.
As we understand it, the main problems to solve for adding linear capabilities to a capability machine like CHERI are related to the move semantics for instructions like \texttt{move}, \texttt{store} and \texttt{load}.
Processor optimizations like pipelining and out-of-order execution rely on being able to accurately predict the registers and memory that an instruction will write to and read from.
Our instructions are a bit clumsy from this point-of-view because, for example, \texttt{move} or \texttt{store} will zero the source register resp. memory location if the value being written is linear.
A solution for this problem could be to add separate instructions for moving, storing and loading linear registers at the cost of additional opcode space.
Adding splice and split will also consume some opcode space.

Another problem is caused by the move semantics for \texttt{load} in the presence of multiple hardware threads.
In this setting, zeroing out the source memory location must happen atomically to avoid race conditions where two hardware threads end up reading the same linear capability to their registers.
This means that a \texttt{load} of a linear capability should behave atomically, similar to a primitive compare-and-swap instruction.
This is in principle not a problem except that atomic instructions are significantly slower than a regular \texttt{load} (on the order of 10x slower or more).
When using \stktokens{}, loads of linear capabilities happen only when a thread has stored its return data capability on the stack and loads it back from there after a return.
Because the stack is a region of memory with very high thread affinity (no other hardware thread should access it, in principle), and which is accessed quite often, well-engineered caching could perhaps reduce the high overhead of atomic loads of linear capabilities.
% If such memory could be (mostly) kept exclusively locked in a cache close to the processor, the overhead of atomic loads in \stktokens{} might be significantly less than \texttt{load}'s worst case.
The processor could perhaps also (be told to) rely on the fact that race conditions should be impossible for loads from linear capabilities (which should in principle be non-aliased) and just use a non-atomic load in that case.

>>>>

With the technical results established, it's worth taking a step back and discuss our work from a moreo high-level perspective.
In the next sections, we specifically give some more thoughts on our use of full abstraction to express WBCF and LSE, and the practical usability of \stktokens{}.

@@

  We have not investigated support for continuations and exceptions in \stktokens{} thoroughly but we expect such support could be added.
  For exceptions, one approach would let callers provide callees with an additional capability for exceptional returns.
  This second capability would be similar to the code part of the return capability pair and signed with the same seal.
  The callee would be able to invoke it to signal that an exception has been thrown after which the caller's code would handle the exception, either by executing an exception handler or by unwinding its stack frame and passing the exception on to its own caller.
  Essentially, this would mean every function would be made responsible for unwinding its own stack frame.
  Continuations are more complicated but could perhaps be treated using similar ideas.
  Alternatively, it might also be possible to have a piece of central trusted code that does the stack unravelling for all stack frames.
  To do this, the trusted code would need to receive a copy of all return seals from the linker.

  % Say something about full-abstractions proofs and feature extensions
  Full abstraction is a property of the whole language.
  In other words, a full abstraction proof must consider all features of a language to make sure that the features don't interact in a way that breaks language abstractions.
  If we have a fully abstract compiler and add a feature to the source or target language, then the new compiler is not necessarily fully abstract anymore.
  Full abstraction would have to be proven for the new compiler to make sure that the new feature does not break existing abstractions in the language.
  Our proof of full-abstraction for \stktokens{} targets a simple capability machine that may not be able to enforce the high-level language abstractions we want, e.g.\ address hiding.
  In other words, the full-abstraction proof cannot be reused immediately.
  However, \stktokens{} is still a good candidate for the enforcement mechanism for well-bracketed control-flow and local-state encapsulation in a real fully abstract compiler.
  Generally speaking, it is worth investigating enforcement mechanisms for full abstraction in a simple setting that allows to quickly try out ideas and verify that the enforcement works.

  % new paragraph: conditional full-abstraction
  Our full-abstraction theorem, Theorem~\ref{thm:full-abstraction} only applies to components that are reasonable and well-formed.
  In other words, if we were to define a compiler phase that targets \srccm{}, then we would also have to show that every program it generates is well-formed and reasonable in order to use the full-abstraction result.
  Without the reasonability constraint, \stktokens{} would have to enforce reasonability instead.
  That is, \stktokens{} would have to dynamically ensure that no return seals or means to obtain them are passed in calls.
  Such checks would impose a performance overhead and are in principle unnecessary for correct code.
  We have instead chosen to rely on the compiler to generate reasonable code that never exhibits the unreasonable behaviour.
  This should be done in a compiler phase where more information about the original program is available, e.g.\ the compilation phase that commits to the low-level translation.
  Similarly for the syntactic constraints given by well-formedness.
  The compiler should make sure to generate code that satisfies the well-formedness judgement, so it can be executed by the machine.

@@

Although this assumption can only be discharged by demonstrating an actual implementation with efficiency measurements, the following notes are based on private discussions with people from the CHERI team as well as our own thoughts on the matter.
As we understand it, the main problems for adding linear capabilities to a capability machine like CHERI are related to the move semantics for instructions like \texttt{move}, \texttt{store} and \texttt{load}.
Processor optimizations like pipelining and out-of-order execution rely on being able to accurately predict the registers and memory that an instruction will write to and read from.
Our instructions are a bit clumsy from this point-of-view because, for example, \texttt{move} or \texttt{store} will zero the source register resp. memory location if the value being written is linear.
A solution for this problem could be to add separate instructions for moving, storing and loading linear registers at the cost of additional opcode space.
Adding splice and split will also consume some opcode space.

Another problem is caused by the move semantics for \texttt{load} in the presence of multiple hardware threads.
In this setting, zeroing out the source memory location must happen atomically to avoid race conditions where two hardware threads end up reading the same linear capability to their registers.
This means that a \texttt{load} of a linear capability should behave atomically, similar to a primitive compare-and-swap instruction.
This is in principle not a problem except that atomic instructions are significantly slower than a regular \texttt{load} (on the order of 10x slower or more).
When using \stktokens{}, loads of linear capabilities happen only when a thread has stored its return data capability on the stack and loads it back from there after a return.
Because the stack is a region of memory with very high thread affinity (no other hardware thread should access it, in principle), and which is accessed quite often, well-engineered caching could perhaps reduce the high overhead of atomic loads of linear capabilities.
% If such memory could be (mostly) kept exclusively locked in a cache close to the processor, the overhead of atomic loads in \stktokens{} might be significantly less than \texttt{load}'s worst case.
The processor could perhaps also (be told to) rely on the fact that race conditions should be impossible for loads from linear capabilities (which should be non-aliased) and just use a non-atomic load in that case.

### 2020-07-24 popl-2019-journal-v/paper.tex
disable todo notes, journal version annotations.
Include a "conflicts of interest declaration" as required by JFP.
Update Lau's affiliation (proposal).
A few final pre-submission modifications.

\definecolor{Blue}{RGB}{92,26,143}
\definecolor{OliveGreen}{RGB}{60,128,49}
\newenvironment{jversion}%{}{}
    {\color{OliveGreen}}{}
\newenvironment{jversionsug}
    {\begin{comment}}{\end{comment}}
\newenvironment{thesisonly}{}{}

\newcommand{\ijversion}[1]%{#1}
    {{\color{OliveGreen} #1}}
\newcommand{\ijversionsug}[1]%{}
    {{\color{Blue} #1}}

@@

\author{Lau Skorstengaard}
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \institution{Aarhus~University}
  \country{Denmark}                    %% \country is recommended
}
\email{lask@cs.au.dk}

@@

\renewcommand\lau[1]{{\color{purple} \sf \footnotesize {LS: #1}}\\}
\renewcommand\dominique[1]{{\color{purple} \sf \footnotesize {DD: #1}}\\}
\renewcommand\lars[1]{{\color{purple} \sf \footnotesize {LB: #1}}\\}

\section{Introduction}
\label{sec:introduction}
\todo[inline]{TODOS ENABLED!!!}
Secure compilation is an active topic of research (e.g.~\citep{devriese_modular_2017,patrignani_hyper_2017,Abate:2018:GCG:3243734.3243745,new_universal_embedding_2016,juglaret_beyond_2016, patrignani_2019}), but a real secure compiler is yet to be built.
Secure compilers preserve source-language (security-relevant) properties even when the compiled code interacts with arbitrary target-language components.
Generally, properties that hold in the source language but not in the target language need to be somehow enforced by the compiler.
Two properties that hold in many high-level source languages are well-bracketed control flow and encapsulation of local state, but they are usually not enforced after compilation to assembly.

@@

  Generally, we have only added material that we believe is valuable to some readers, and we have worked hard to explain the more technical material and make it digestible.
  Additionally, while Sections~\ref{sec:cap-mach-w-seal-and-lin}, \ref{sec:stktokens-explained},~\ref{sec:form-secur-with},~\ref{sec:discussion} and~\ref{sec:related-work} are intended for all readers, we have kept the more technical sections about the proof of full abstraction separate in Sections~\ref{sec:fa-proof} and~\ref{sec:rec-dom-eq}, so that it can be easily skipped by readers who prefer to skip those explanations.
\end{jversion}
% Notes on new contributions (based on the journal version environment):
% More details about the definition of the cap. machine (decoding/encoding functions)
% Well-formedness definitions
% Reasonability definition
% Expanded discussion section
% More detailed description of worlds and
%   the logical relation used

@@

Plenty of other papers enforce well-bracketed control flow at a low level but most are restricted to preventing particular types of attacks and enforce only partial correctness of control flow.
This includes particularly the line of work on control-flow integrity~\citep{abadi_control-flow_2005}.
This technique prevents certain classes of attacks by sanitizing addresses before direct and indirect jumps based on static control graph information and a protected shadow stack.
Contrary to \stktokens{}, CFI can be implemented on commodity hardware rather than capability machines.
However, its attacker model is different, and its security goals are weaker.
They assume an attacker that is unable to execute code but can overwrite arbitrary data at any time during execution (to model buffer overflows).
In terms of security goals, the technique does not enforce local stack encapsulation.
Also, it only enforces a weak form of control flow correctness saying that jumps stay within the program's static control flow graph~\cite{Abadi2005Theory}.
Such a property ignores temporal properties and seems hard to use for reasoning.
There is also more and more evidence that these partial security properties are not enough to prevent realistic attacks in practice~\citep{Evans:2015:CJW:2810103.2813646,Carlini2015ControlFlowBending}.

@@

Local capabilities can be used to ensure well-bracketed control-flow and local-state encapsulation as demonstrated by \citet{skorstengaard_reasoning_2017}.
Recently, \citet{tsampas_2019} demonstrated that an extension of local capabilities with multiple linearly ordered colours can be used to enforce the life time of stack references.
Specifically, a stack reference should not be able to outlive the stack frame it points to.
If \stktokens{} was extended with stack references, then it would also enforce reference life times.
Specifically in order to return from a call, we must use the return token, i.e.\ the stack.
The stack is linear, so if there are references to it, aside from the stack capability itself, then we cannot have a complete return token.
This means that we have to splice all the stack references together with the stack capability to complete the return token in order to return.
\citet{tsampas_2019} allow (almost) normal references that can be stored in multiple places at the same time.
This means that their approach is more like C than what \stktokens{} has to offer.
As explained in Section~\ref{sec:introduction}, these approaches have the downside that they require stack clearing (including unused parts) on boundary crossings.

@@

\begin{acks}
  This research was supported in part by the ModuRes Sapere Aude Advanced Grant
  from The Danish Council for Independent Research for the Natural Sciences (FNU).
  Support for an STSM was received from COST Action EUTypes (CA15123).
  Dominique Devriese held a Postdoctoral fellowship from the Research Foundation
  Flanders (FWO) during most of this research.
  This research was supported in part by the Research Foundation Flanders (FWO).
\end{acks}

>>>>

\definecolor{Blue}{RGB}{92,26,143}
\definecolor{OliveGreen}{RGB}{60,128,49}
\newenvironment{jversion}{}{}
    % {\color{OliveGreen}}{}
\newenvironment{jversionsug}
    {\begin{comment}}{\end{comment}}
\newenvironment{thesisonly}{}{}

\newcommand{\ijversion}[1]{#1}
    % {{\color{OliveGreen} #1}}
\newcommand{\ijversionsug}[1]{}
    % {{\color{Blue} #1}}

@@

\author{Lau Skorstengaard}
% \orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \institution{Toitware}
  \country{Denmark}                    %% \country is recommended
}
\authornote{Research performed while the author was working for Aarhus University.}
\email{l030689@gmail.com}

@@

% \renewcommand\lau[1]{{\color{purple} \sf \footnotesize {LS: #1}}\\}
% \renewcommand\dominique[1]{{\color{purple} \sf \footnotesize {DD: #1}}\\}
% \renewcommand\lars[1]{{\color{purple} \sf \footnotesize {LB: #1}}\\}
\renewcommand\lau[1]{}
\renewcommand\dominique[1]{}
\renewcommand\lars[1]{}

\section{Introduction}
\label{sec:introduction}
\todo[inline]{TODOS ENABLED!!!}
Secure compilation is an active topic of research (e.g.~\citep{devriese_modular_2017,patrignani_hyper_2017,Abate:2018:GCG:3243734.3243745,new_universal_embedding_2016,juglaret_beyond_2016, patrignani_2019,barthe_formal_2019}), but real secure compilers are still a rare sight.
Secure compilers preserve source-language (security-relevant) properties even when the compiled code interacts with arbitrary target-language components.
Generally, properties that hold in the source language but not in the target language need to be somehow enforced by the compiler.
Two properties that hold in many high-level source languages are well-bracketed control flow and encapsulation of local state, but they are usually not enforced after compilation to assembly.

@@

  Generally, we have only added material that we believe is valuable to some readers, and we have worked hard to explain the more technical material and make it digestible.
  Additionally, while Sections~\ref{sec:cap-mach-w-seal-and-lin}, \ref{sec:stktokens-explained},~\ref{sec:form-secur-with},~\ref{sec:discussion} and~\ref{sec:related-work} are intended for all readers, we have kept the more technical sections about the proof of full abstraction separate in Sections~\ref{sec:fa-proof} and particularly \ref{sec:rec-dom-eq}, so that it can be easily skipped by readers who prefer to do so.
\end{jversion}
% Notes on new contributions (based on the journal version environment):
% More details about the definition of the cap. machine (decoding/encoding functions)
% Well-formedness definitions
% Reasonability definition
% Expanded discussion section
% More detailed description of worlds and
%   the logical relation used

@@

Plenty of other papers enforce well-bracketed control flow at a low level but most are restricted to preventing particular types of attacks and enforce only partial correctness of control flow.
This includes particularly the line of work on control-flow integrity~\citep{abadi_control-flow_2005}.
This technique prevents certain classes of attacks by sanitizing addresses before direct and indirect jumps based on static control graph information and a protected shadow stack.
Contrary to \stktokens{}, CFI can be implemented on commodity hardware rather than capability machines.
However, its attacker model is different, and its security goals are weaker.
They assume an attacker that is unable to execute code but can overwrite arbitrary data at any time during execution (to model buffer overflows).
In terms of security goals, the technique does not enforce local stack encapsulation.
Also, it only enforces a weak form of control flow correctness saying that jumps stay within the program's static control flow graph~\cite{Abadi2005Theory}.
Such a property ignores temporal properties and seems hard to use for reasoning.
There is also more and more evidence that these partial security properties are not enough to prevent realistic attacks in practice~\citep{Evans:2015:CJW:2810103.2813646,Carlini2015ControlFlowBending,van_der_veen_dynamics_2017}.

@@

Local capabilities can be used to ensure well-bracketed control-flow and local-state encapsulation as demonstrated by \citet{skorstengaard_reasoning_2017}.
Recently, \citet{tsampas_2019} demonstrated that an extension of local capabilities with multiple linearly ordered colours can be used to enforce the life time of stack references.
Specifically, a stack reference should not be able to outlive the stack frame it points to.
If \stktokens{} was extended with stack references, then it would also enforce reference life times.
Specifically in order to return from a call, we must use the return token, i.e.\ the stack.
The stack is linear, so if there are references to it, aside from the stack capability itself, then we cannot have a complete return token.
This means that we have to splice all the stack references together with the stack capability to complete the return token in order to return.
\citet{tsampas_2019} allow (almost) normal references that can be stored in multiple places at the same time.
This means that their approach is more like C than \stktokens{}.
As explained in Section~\ref{sec:introduction}, these approaches have the downside that they require stack clearing (including unused parts) on boundary crossings.

@@

\begin{acks}
  This research was supported in part by the ModuRes Sapere Aude Advanced Grant
  from The Danish Council for Independent Research for the Natural Sciences (FNU).
  Support for an STSM was received from COST Action EUTypes (CA15123).
  Dominique Devriese held a Postdoctoral fellowship from the Research Foundation
  Flanders (FWO) during most of this research.
  This research was supported in part by the Research Foundation Flanders (FWO).
  Conflicts of Interest: None.
\end{acks}
